Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-22 pkg/disttree/inst/draft/draft.Rnw
further improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2069 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: aa76bab9089ba006676822a08f8e104a7a08ab11

transpgrey <- rgb(0.190,0.190,0.190, alpha = 0.2)

@@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

@@

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgrey", type = "l", lty = 2)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgrey", type = "l", lty = 2)

@@

A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$. Based on the
corresponding probability density function $f(Y; \bm{\theta})$ the log-likelihood
function is defined $\ell(\bm{\theta}; Y) = \log\{f(Y; \bm{\theta})\}$. The GAMLSS
framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
distributional families with parameterizations corresponding to location, scale,
and shape. Furthermore, censoring and/or truncation of these distributions
can be incorporated in the usual straightforward way \fixme{reference needed}.

@@

In a first step, we show a detailed comparison of the competing models
for one observation station, Axams in Tyrol (selected somewhat arbitrarily
as the closest in the data set to Innsbruck, capital of Tyrol that three of the
authors are based in). As for all other stations, daily precipitation observations and
numerical weather predictions are available for the month of July from 1985
to 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
forecasts from the distributional forest, trained on 1985--2008, for July 24
in 2009--2012 have already been shown as a motivational example. The figure
depicts the point masses at zero (i.e., the probability of a dry day) and
the forecasted probability density function for the total amount of precipitation,
showing that these differ considerably in location, scale, and the amount of
censoring while conforming quite well with the actual precipitation observations
from those days. But, of course, this is just one selected day and we are interested
in assessing how well the distributional fits are calibrated and what the
relative predictive performance is.

@@

Finally, it is of interest how this improvement in predictive performance
by the distributional forest is accomplished, i.e., which of the 80~covariates
are selected in the trees of the forest. As the 100~trees of the forest

\fixme{still not finished}

As shown in Table~\ref{tab:models} the distributional forest 
includes all of the available covariates. A single tree model would allow 
to easily analyze which covariates are used as split variables in the nodes 
and hence have a significant impact on the response variable. In a forest 
model though, for each node in each of the trees a subset of the covariates 
is chosen randomly as possible split variables to select one from. Moreover, 
each tree is built on a resampled version of the learning data. For these 
reasons the choices of split variables vary over the trees of a forest making 
it difficult to analyze the impact of covariates just by investigating the 
different trees.

To get an overview of which covariates have the most influential 
effect on the response variable a standard measurement of variable importance 
for forest models has been applied. Based on the mean decrease in accuracy 
regarding the CRPS the 10 leading covariates of the above described distributional 
forest for station Axams learned on the first 24 years are listed in 
Figure~\ref{fig:axams-varimp}. As expected, the numerical forecast for total precipitation
(\emph{tp}) is the most influential variable. In particular, five variations of 
this variable are among the top ten of the list of covariates ordered by 
variable importance.

@@

Until now the focus has been on one observation station only. To test and 
compare the four methods under different geographical circumstances they
are applied on all 95 stations. For each station, the first 
24 years are used as learning data set and predictions are made for the 
ensuing 4 years of available observations. For these out-of-sample predictions 
the CRPS skill score with the EMOS model as reference model is calculated 
for each method at each station. In Figure~\ref{fig:all-crps} the boxplots 
of the CRPS skill score values are plotted illustrating that the distributional 
forest leads to the highest improvement averaged over all stations.
The green line represents the results for the station Axams.
%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))
%<<eval = TRUE, echo = FALSE>>=
<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("rain_pred_24to4.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
  load("rain_pred_24to4.rda")
} else {

@@

\begin{figure}[t!]
\centering
<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, width = 7>>=
  matplot(t(s[,]), type = "l", lwd = 2, 
          col = gray(0.5, alpha = 0.2),
          lty = 1, axes = FALSE, 
          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
lines(s[70,], col = "limegreen", type = "o", pch = 19, lwd = 2)  
# Station Axams is the 77th station which is the 70th in the list of complete stations
  boxplot(s, add = TRUE, col = "transparent")
  abline(h = 0, col = pal["EMOS"], lwd = 2)

@
\caption{\label{fig:all-crps}CRPS skill score for each station (grey lines) 
and aggregated over all stations (boxplots). Station Axams is highlighted in green 
and the horizontal orange line pertains to the reference model EMOS. The models are 
learned on data from 24 years (1985--2008) and validated on 4 years (2009--2012).}
\end{figure}

Figure~\ref{fig:all-map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
Data, \cite{srtm}) with the location of all observation stations included in the 
data set. Each station is illustrated by a symbol with the type of symbol 
representing the method that performed best in terms of CRPS at this station. 
The color of the symbol indicates the difference between the CRPS of the 
distributional forest and the CRPS of the best performing among the other 
three models (prespecified GAMLSS, boosted GAMLSS and EMOS).
Even though for some stations, especially in East Tyrol, one of the two GAMLSS
models and for a few others the EMOS model leads to the best results, for the 
majority of stations the distributional forest performs best.

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.99\textwidth}
<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=

@@

@
\caption{\label{fig:all-map}Map of Tyrol showing which model performed best at 
which station (type of symbol). The color codes whether the distributional 
forest had lower or higher CRPS (red or blue) compared to the best of the 
other three models (prespecified GAMLSS, boosted GAMLSS and EMOS).}
\end{figure}

@@

From the results in Section~\ref{sec:precipitation} we can conclude 
that distributional forests provide a powerful alternative framework requiring 
only little effort to set up a model which is competitive to commonly used 
methods or even outperforms them. Even though a proper selection of covariates 
and specification of interactions in advance can lead to better fitting 
(and probably less complex) models, this presupposes profound knowledge of 
all influencing factors. Otherwise, important aspects might be missed or 
effects which do not correspond to the truth might be imposed on the model. 
This risk can be avoided by applying a boosting algorithm for an automatic 
variable selection. In this case the trade-off is often a remarkably higher 
computational effort. To build a distributional forest the user is not 
required to provide any expert knowledge or additional information about 
the setting as variable selection and detection of interactions are done 
automatically by the tree-structure of the model. This feature is 
particularly advantageous in case of a very high number of covariates. 
Therefore, the above discussed application illustrates the strengths of 
distributional forests as they offer a very flexible framework which 
enables an easy use of the method and at the same time lead to an overall 
improvement in the predictive performance in this setting. Especially 
in situations where no or only little information is available about 
possible influences, effects or interactions this novel methodology 
can be seen as a good and recommendable compromise being able to deal 
with a wide range of modeling situations in a reasonable way.

%Even though many situations demand for the use of specifically tailored 
%models providing the needed features and therefore being clearly advantageous 
%and the best choice, it is often hard to tell in advance which features 
%will be required. In these cases a distributional forest is a 
%recommendable compromise providing reasonable models and performing 
%well without any prespecification or additional knowledge needed in advance.

\section*{Computational details}
The corresponding implementation to the novel methods presented 
in this article is available on \textsf{R}-Forge in an \textsf{R} 
package called \textbf{disttree} within the project \textbf{partykit} 
(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
\code{distfit} fits a distributional model and is called within the
tree-building function \code{disttree} and the forest-building
function \code{distforest}. 
Within \code{disttree} either the \code{mob} or \code{ctree} function 
can be applied as a framework to build a tree. Similarly, the function 
\code{distforest} is based on the function \code{cforest}. All three 
functions, \code{mob}, \code{ctree} and \code{cforest}, are provided 
in the package \textbf{partykit} (\citealt{Hothorn+Zeileis:2015}). 
To specify a distribution family the user can hand over one of the 
following objects as input argument in \code{distfit}, \code{disttree} 
and \code{distforest}: 
\begin{itemize}
\item a GAMLSS family object from the \textsf{R} package \textbf{gamlss.dist} 
(\citealt{Stasinopoulos+Rigby:2007}).
\item a family list generating function from the package \textbf{disttree}.
\item a complete family list from the package \textbf{disttree}.
\item a custom list containing all required information about the 
distribution family.
\end{itemize}

In Section~\ref{sec:precipitation} the prespecified GAMLSS were 
built using the function \code{gamlss} from the \textsf{R} package 
\textbf{gamlss} (\citealt{Stasinopoulos+Rigby:2007}). For the 
boosted GAMLSS the function \code{gamboostLSS} from the \textsf{R} 
package \textbf{gamboostLSS} (\citealt{Hofner+Mayr+Schmid:2016}) 
was applied. The EMOS models were built using the function 
\code{crch} from the \textsf{R} package \textbf{crch} 
(\citealt{Messner+Mayr+Zeileis:2016})

>>>>

transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

@@

<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = ytree, col = pal["forest"], lwd=7)
@
\endminipage};
\node (b) at (5,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@

<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt="n", yaxt="n", ann=FALSE, col = "slategray", pch = 19)
box(lwd=5)
#lines(x = x, y = ytrue, col = "gray", lwd=5, main = "")
lines(x = x, y = yforest, col = pal["forest"], lwd=7, main = "")
@
\endminipage};
\node (c) at (10,0) 
{
\minipage{0.29\textwidth}
\centering
<<motivation_distforest, echo=FALSE, results=hide>>=
## distforest
set.seed(4723)
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
xt <- 2*(x-0.5)
var <- exp(-(2*xt)^4)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
x <- c(1:nobs)/nobs
@

@@

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2)

@@

A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$. Based on the
corresponding probability density function $f(Y; \bm{\theta})$ the log-likelihood
function is defined $\ell(\bm{\theta}; Y) = \log\{f(Y; \bm{\theta})\}$. The GAMLSS
framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
distributional families with parameterizations corresponding to location, scale,
and shape. Furthermore, censoring and/or truncation of these distributions
can be incorporated in the usual straightforward way
\citep[see e.g.,][Chapter~7.2]{Long:1997}.

@@

In a first step, we show a detailed comparison of the competing models
for one observation station, Axams in Tyrol (selected somewhat arbitrarily
as the closest in the data set to Innsbruck, capital of Tyrol and base of three of the
authors). As for all other stations, daily precipitation observations and
numerical weather predictions are available for the month of July from 1985
to 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
forecasts from the distributional forest, trained on 1985--2008, for July 24
in 2009--2012 have already been shown as a motivational example. The figure
depicts the point masses at zero (i.e., the probability of a dry day) and
the forecasted probability density function for the total amount of precipitation,
showing that these differ considerably in location, scale, and the amount of
censoring while conforming quite well with the actual precipitation observations
from those days. But, of course, this is just one selected day and we are interested
in assessing how well the distributional fits are calibrated and what the
relative predictive performance is.

@@

Finally, it is of interest how this improvement in predictive performance
by the distributional forest is accomplished, i.e., which of the 80~covariates
are selected in the trees of the forest. As the 100~trees of the forest
do not allow to simply assess the variables' role graphically, a common
solution for random forests in general is to consider variable importance
measures. Here, this is defined as the amount of change in CRPS when
the association between one covariate and the response variable is
artificially broken through permutation (and thus also breaking the
association to the remaining covariates).

Figure~\ref{fig:axams-varimp} shows 10~covariates with the highest permutation
importance (i.e., change in CRPS) for station Axams. As expected the NWP outputs
for total precipitation (\emph{tp}) are particularly important along with total
column-integrated condensate (\emph{tcolc}). Also, both variables occur
in various transformations such as means (either of the full day or certain parts
of the afternoon), spreads, or minima/maxima. Thus, while the covariates
themselves are not surprising, a specification with all the transformations
would typically not have been considered in a GAMLSS.

@@

<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("rain_pred_24to4.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
  load("rain_pred_24to4.rda")
} else {

@@

\begin{figure}[p!]
\centering
<<rain_all_crps_skill_score, fig=TRUE, echo=FALSE, height = 4.5, width = 7>>=
par(mar = c(3.1, 4.1, 1.1, 2.1))
matplot(t(s[,]), type = "l", lwd = 2, 
          col = gray(0.5, alpha = 0.2),
          lty = 1, axes = FALSE, 
          xlab = "", ylab = "CRPS skill score", xlim = c(0.5, 3.5))
lines(s[70,], col = "limegreen", type = "o", pch = 19, lwd = 2)  
# Station Axams is the 77th station which is the 70th in the list of complete stations
  boxplot(s, add = TRUE, col = "transparent")
  abline(h = 0, col = pal["EMOS"], lwd = 2)

@
\caption{\label{fig:all-crps}CRPS skill score for each station (gray lines
with boxplots superimposed). Station Axams is highlighted in green 
and the horizontal orange line pertains to the reference model EMOS. The models are 
learned on 1985--2008 and validated for 2009--2012.}

\vspace*{0.4cm}

\setkeys{Gin}{width=0.99\textwidth}
<<map, fig=TRUE, echo=FALSE, height=6.1, width=9>>=

@@

@
\caption{\label{fig:all-map}Map of Tyrol coding the best-performing model
for each station (type of symbol). The color codes whether the distributional 
forest had higher (green) or lower (red) CRPS compared to the best of the 
other three models.}
\end{figure}

After considering only one observation station up to now, a second step
evaluates and compares the competing methods on all 95~stations. As in the previous
section, all models are learned on the first 24~years and evaluated by the
average CRPS on the last 4~years. More specifically, the CRPS skill score
against the EMOS model is computed for the out-of-sample predictions at each
station and visualized by parallel coordinates plots with boxplots superimposed
in Figure~\ref{fig:all-crps}. Overall distributional forests have a slightly
higher improvement in CRPSS compared to the two GAMLSS which is best seen
by looking at the boxplots and also in the green line representating the results
for station Axams. The underlying parallel coordinates additionally
bring out that the prespecified GAMLSS sometimes performs rather differently
(sometimes better, sometimes worse) compared to the two data-driven models.
Finally, because some CRPSS become negative, it is also clear that the simple
EMOS performs better than the other models for a few of the stations.

%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))
%<<eval = TRUE, echo = FALSE>>=

To assess whether these differences in predictive performance are due to differences
in the topology, Figure~\ref{fig:all-map} shows a brief summary for each station
on a map of Tyrol (based on the SRTM 90m Digital Elevation 
Data, \cite{srtm}). Each station is illustrated by a symbol that conveys which
model performed best in terms of CRPS on the last 4~years of the data.
Additionally, the color of the symbol indicates the CRPS difference between 
distributional forest and the best-performing other model. Green signals that
the distributional forest performs better than the other models whereas red
signals that another model performs better. Overall the distributional forest
performs on par (gray) or better (green) for the majority of stations. Only
for a few stations in the north-east EMOS performs best, and in East Tyrol
the prespecified EMOS performs particularly well.

@@

Distributional regression modeling is combined with tree-based modeling
to obtain a novel and flexible method for probabilistic forecasting.
The resulting distributional trees and forests can capture abrupt and
nonlinear effects and interactions in a data-driven way. By basing the
split point and split variable selection on a full likelihood and
corresponding score function, the trees and forests can not only pick
up changes in the location but also the scale or shape of a distributional
family.

Distributional forests are an attractive alternative when prespecifying or boosting
all effects and interactions in a GAMLSS model is challenging. They are
rather straightforward to specify requiring only little prior subject
matter knowledge and also work well in the presence of many potential
covariates. The application to precipitation forecasting in complex terrain
illustrates that distributional forests often perform on par or even
better than their GAMLSS counterparts. Hence, they form a useful addition
to the toolbox of probabilistic forecasts, especially in meteorology.

\section*{Computational details}

The proposed methods are in the \textsf{R} package \textbf{disttree} on
\textsf{R}-Forge within the \textbf{partykit} project
(\url{https://R-Forge.R-project.org/projects/partykit/}). The function
\code{distfit} fits distributional models by maximum likelihood, which is
used as the basis for the tree-building function \code{disttree}, upon which
the \code{distforest} is built. All functions can either be used with
GAMLSS family objects from the \textsf{R} package \textbf{gamlss.dist} 
(\citealt{Stasinopoulos+Rigby:2007}) or with custom lists containing
all required information about the distribution family.

In addition to \textbf{disttree}, Section~\ref{sec:precipitation} employs
the functions \code{crch} from the \textsf{R} package \textbf{crch} 
\citep{Messner+Mayr+Zeileis:2016} for the EMOS models,
\code{gamlss} from the \textsf{R} package \textbf{gamlss}
\citep{Stasinopoulos+Rigby:2007} for the prespecified GAMLSS, and
\code{gamboostLSS} from the \textsf{R} package \textbf{gamboostLSS}
\citep{Hofner+Mayr+Schmid:2016} for the boosted GAMLSS.

\fixme{replication code?}
